Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data

### 2021-06-29 paper/paper.Rmd
Start on second revision

Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution. This affects visualizations made on low-dimensional projections, like the tour [@As85,@BCAH05]. The effect is described by @diaconis1984, that most low-dimensional linear projections are approximately Gaussian, with observations concentrating in the center. This has motivated the development of indexes for projection pursuit which search for departure from normality. It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976,@10.1093/biomet/asp084]: all observations can collapse into a single point.  These issues also persist with non-linear dimension reduction techniques, and are often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.

\textcolor{red}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour. Two-dimensional linear projections of data, generated by sampling uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$), are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection. \textcolor{red}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}

@@

\textcolor{red}{Common solutions to overplotting include transparency and binned or density displays} [@chang]. \textcolor{red}{These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.}

\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions. This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data. When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a uniform distribution in the low-dimensional display. In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume. The observed projected distribution is representative of the original high-dimensional distribution in this sense.}

@@

To understand how the $p$-dimensional volume is projected onto a $2$-dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius. This dependence was described in @Laa:2020wkm. We start from a $p$-dimensional hypersphere, with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r; p, R)$, where $r$ can be any radius within $[0, R]$. The relative projected volume is then given as the ratio of these two quantities,
\begin{equation}
v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation}
This ratio is of particular interest because it gives the 2-dimensional radial cumulative distribution function (CDF) of points when assuming a uniform distribution within the $p$-dimensional hypersphere.

@@

The first transformation is to replace $r_y$ with $v_{2} (r_y; p, R)$. Since this is the radial CDF of the assumed underlying distribution, we expect that $v_{2} (r_y; p, R)$ is approximately uniformly distributed in radius. We then transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk. This inverse is defined via
\begin{equation}
v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y
\end{equation}
and thus
\begin{equation}
v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}.
\end{equation}

@@

```{r circles, fig.cap="Equidistant concentric circles, for $p=2, 3, 10, 100$, illustrating the radial transformation. The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.", out.width="100%", fig.align = "center", fig.height=2}

@@

@@

\textcolor{red}{Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data. In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions. This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation. Scaling is commonly also combined with centering as a preprocessing step. This is less important in our implementation because we always center the projected data before applying our transformation (i.e. the projected mean defines the origin of our polar coordinate system in the plane). Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.}

\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of a variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}

\textcolor{red}{The third common pre-processing would involve a dimension reduction step. Even with a relatively low number of dimensions, e.g. with $p=10$, we see noticeable piling. However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise. In this case it is advisable to first reduce the number of dimensions, for example using principal components analysis. Other dimension reduction techniques could be used. The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two. }

@@

\noindent and uses `gam` to set the $\gamma$ parameter for computing $p_{\mathrm{eff}}$, and the overall range `R` used for trimming. Both these parameters are described in Section \ref{sec:params} above. Finally, `half_range` sets the scale parameter $s$ to adjust the scale to the drawing canvas. The ratio $R/s$ sets the scale for fitting the displayed data on the plotting canvas, by default $s = R$ and we apply the scaling factor $0.9$ to contain the projected points within the display. When adjusting $R$ the user should take care to adjust $s$ accordingly.

@@

An alternative is the slice tour [@sliceTour] which  allows distributions
of points around the center of the data to be explored using sections instead of projections.
\textcolor{red}{Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.}
The slice tour is useful when there are large numbers of observations or if there is concave structure in the data.
\textcolor{red}{Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation. In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.}

>>>>

Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution. This affects visualizations made on low-dimensional projections, like the tour [@As85;@BCAH05]. The effect is described by @diaconis1984, that most low-dimensional linear projections are approximately Gaussian, with observations concentrating in the center. This has motivated the development of indexes for projection pursuit which search for departure from normality. It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976;@10.1093/biomet/asp084]: all observations can collapse into a single point.  These issues also persist with non-linear dimension reduction techniques, and are often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.

Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space. Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour. Two-dimensional linear projections of data, generated by sampling uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$), are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection. (Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)

@@

Common solutions to overplotting include transparency and binned or density displays [@chang]. These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.

The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions. This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data. When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a uniform distribution in the low-dimensional display. In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume. The observed projected distribution is representative of the original high-dimensional distribution in this sense.

@@

To understand how the $p$-dimensional volume is projected onto a $2$-dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius. This dependence was described in @Laa:2020wkm. We start from a $p$-dimensional hypersphere, with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r; p, R)$, where $r$ can be any radius within $[0, R]$. The relative projected volume is then given as the ratio of these two quantities,
\begin{equation}
v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation}

@@

\textcolor{red}{We transform $r_y$ in two steps. First, we replace it with $v_{2} (r_y; p, R)$ to redistribute the volume uniformly by the radius. The second step is then to } 
transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk. This inverse is defined via $v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y$ and thus $v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}$.

@@

```{r circles, fig.cap="Concentric circles that are equidistant before the application of the radial transformation, for $p=2, 3, 10, 100$, illustrating the effect of the transformation with increasing number of dimensions. The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.", out.width="100%", fig.align = "center", fig.height=2}

@@

\textcolor{red}{At this point a few characteristics of the transformation should be noted. First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph. In addition, since only the radial component is transformed, angles between the points will not be altered. (This can also be shown explicitly, from mathematical reformulation of the angles between two points after the transformation.) Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$). For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space. Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.}

@@

Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data. In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions. This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation. Scaling is commonly also combined with centering as a preprocessing step. This is less important in our implementation because we always center the projected data before applying our transformation (i.e. the projected mean defines the origin of our polar coordinate system in the plane). Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.

We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of a variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.

The third common pre-processing would involve a dimension reduction step. Even with a relatively low number of dimensions, e.g. with $p=10$, we see noticeable piling. However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise. In this case it is advisable to first reduce the number of dimensions, for example using principal components analysis. Other dimension reduction techniques could be used. The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two.

@@

\noindent and uses `gam` to set the $\gamma$ parameter for computing $p_{\mathrm{eff}}$, and the overall range `R` used for trimming. Both these parameters are described in Section \ref{sec:params} above.

For consistency with other display functions in the `tourr` package we also pass in the scale parameter $s$ via the `half_range` argument. By default this is evaluated from the range of the data, but kept as a free parameter to zoom in (without any transformation) if necessary. This can be important, e.g. when the range is large because of an outlying point. Note here that
the ratio $R/s$ sets the scale for fitting the displayed data on the plotting canvas, by default $s = R$. When adjusting $R$ the user should take care to adjust $s$ accordingly.

@@

\textcolor{red}{Note here that adjusting the resolution using the burning sage transformation will lead to an overall decrease in densitiy near the center. When studying the separation between groups it is important to keep in mind that the effect is the same within and between clusters. Thus, if points within a cluster remain grouped while separation between clusters becomes more apparant, this can be interpreted as a true finding. Conversely, in the case of a visual artifact we expect a general reduction of density that does not lead to better separation between the clusters.}

An alternative is the slice tour [@sliceTour] which  allows distributions
of points around the center of the data to be explored using sections instead of projections.
Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.
The slice tour is useful when there are large numbers of observations or if there is concave structure in the data.
Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation. In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.

### 2021-06-29 paper/paper.tex
Start on second revision

Paradoxically, the curse of dimensionality inverts for dimension
reduction, resulting in an excessive amount of observations near the
center of the distribution. This affects visualizations made on
low-dimensional projections, like the tour \citep[\citet{BCAH05}]{As85}.
The effect is described by \citet{diaconis1984}, that most
low-dimensional linear projections are approximately Gaussian, with
observations concentrating in the center. This has motivated the
development of indexes for projection pursuit which search for departure
from normality. It is also related to what is called ``data piling'' in
high-dimension low-sample size data
\citep[\citet{10.1093/biomet/asp084}]{10.2307/27639976}: all
observations can collapse into a single point. These issues also persist
with non-linear dimension reduction techniques, and are often referred
to as the ``crowding problem'', which methods like t-Distributed
Stochastic Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.

\textcolor{red}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.}
Figure \ref{fig:density} illustrates this crowding problem, as would be
experienced when using a tour. Two-dimensional linear projections of
data, generated by sampling uniformly within \(p\)-dimensional
hyperspheres (\(p=3, 10, 100\)), are displayed as hexbin plots. Color
indicates log count of the bin, with yellow mapping the highest counts.
As \(p\) increases the density concentrates in the center of the
projection.
\textcolor{red}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}

@@

\textcolor{red}{Common solutions to overplotting include transparency and binned or density displays}
\citep{chang}.
\textcolor{red}{These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.}

\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions. This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data. When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a uniform distribution in the low-dimensional display. In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume. The observed projected distribution is representative of the original high-dimensional distribution in this sense.}

@@

To understand how the \(p\)-dimensional volume is projected onto a
\(2\)-dimensional plane, we study what fraction of the total volume is
projected onto the area of a disk depending on its radius. This
dependence was described in \citet{Laa:2020wkm}. We start from a
\(p\)-dimensional hypersphere, with radius \(R\) and volume \(V(R, p)\),
and its projected volume onto a centered \(2\)-dimensional disk of
radius \(r\), \(V_{2D}(r; p, R)\), where \(r\) can be any radius within
\([0, R]\). The relative projected volume is then given as the ratio of
these two quantities, \begin{equation}
v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation} This ratio is of particular interest because it gives the
2-dimensional radial cumulative distribution function (CDF) of points
when assuming a uniform distribution within the \(p\)-dimensional
hypersphere.

@@

The first transformation is to replace \(r_y\) with
\(v_{2} (r_y; p, R)\). Since this is the radial CDF of the assumed
underlying distribution, we expect that \(v_{2} (r_y; p, R)\) is
approximately uniformly distributed in radius. We then transform
\(v_{2} (r_y; p, R)\) using the inverse of \(v_{2} (r_y; 2, R)\), to go
from a uniform distribution in radius to a uniform distribution in area
of the disk. This inverse is defined via \begin{equation}
v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y
\end{equation} and thus \begin{equation}
v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}.
\end{equation}

@@

\caption{Equidistant concentric circles, for $p=2, 3, 10, 100$, illustrating the radial transformation. The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.}\label{fig:circles}
\end{figure}

@@

\textcolor{red}{Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data. In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions. This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation. Scaling is commonly also combined with centering as a preprocessing step. This is less important in our implementation because we always center the projected data before applying our transformation (i.e. the projected mean defines the origin of our polar coordinate system in the plane). Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.}

\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of a variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}

\textcolor{red}{The third common pre-processing would involve a dimension reduction step. Even with a relatively low number of dimensions, e.g. with $p=10$, we see noticeable piling. However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise. In this case it is advisable to first reduce the number of dimensions, for example using principal components analysis. Other dimension reduction techniques could be used. The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two. }

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tourr}\OperatorTok{::}\KeywordTok{animate}\NormalTok{(}
\NormalTok{  data,}
  \DataTypeTok{tour_path =}\NormalTok{ tourr}\OperatorTok{::}\KeywordTok{grand_tour}\NormalTok{(),}
  \DataTypeTok{display =} \KeywordTok{display_sage}\NormalTok{(gam, R, half_range)}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

\noindent and uses \texttt{gam} to set the \(\gamma\) parameter for
computing \(p_{\mathrm{eff}}\), and the overall range \texttt{R} used
for trimming. Both these parameters are described in Section
\ref{sec:params} above. Finally, \texttt{half\_range} sets the scale
parameter \(s\) to adjust the scale to the drawing canvas. The ratio
\(R/s\) sets the scale for fitting the displayed data on the plotting
canvas, by default \(s = R\) and we apply the scaling factor \(0.9\) to
contain the projected points within the display. When adjusting \(R\)
the user should take care to adjust \(s\) accordingly.

@@

An alternative is the slice tour \citep{sliceTour} which allows
distributions of points around the center of the data to be explored
using sections instead of projections.
\textcolor{red}{Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.}
The slice tour is useful when there are large numbers of observations or
if there is concave structure in the data.
\textcolor{red}{Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation. In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.}

>>>>

Paradoxically, the curse of dimensionality inverts for dimension
reduction, resulting in an excessive amount of observations near the
center of the distribution. This affects visualizations made on
low-dimensional projections, like the tour \citep{As85, BCAH05}. The
effect is described by \citet{diaconis1984}, that most low-dimensional
linear projections are approximately Gaussian, with observations
concentrating in the center. This has motivated the development of
indexes for projection pursuit which search for departure from
normality. It is also related to what is called ``data piling'' in
high-dimension low-sample size data
\citep{10.2307/27639976, 10.1093/biomet/asp084}: all observations can
collapse into a single point. These issues also persist with non-linear
dimension reduction techniques, and are often referred to as the
``crowding problem'', which methods like t-Distributed Stochastic
Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.

Here we are providing a solution to fix a long-standing problem in the
tour, where high-dimensional data is viewed by a movie of
low-dimensional projections. Points crowding the center of the drawing
canvas detract from reading the plots, even with relatively few
dimensions. This is a structural problem related purely to the geometry
of projections from high-dimensional spaces. We show that it can be
alleviated with a radial transformation on the low-dimensional space.
Figure \ref{fig:density} illustrates this crowding problem, as would be
experienced when using a tour. Two-dimensional linear projections of
data, generated by sampling uniformly within \(p\)-dimensional
hyperspheres (\(p=3, 10, 100\)), are displayed as hexbin plots. Color
indicates log count of the bin, with yellow mapping the highest counts.
As \(p\) increases the density concentrates in the center of the
projection. (Note that, sampling from a uniform in a sphere is used for
this illustration for several reasons. Any linear projection is
equivalent because of the rotation invariance of the distribution. The
uniform sampling places focus on the space, points are spread evenly
across the volume, and not a data distribution. Also, \(p=10\) would
already be considered high-dimensional, as seen by the piling observed
in the middle plot.)

@@

Common solutions to overplotting include transparency and binned or
density displays \citep{chang}. These are designed to tackle large
sample size, rather than dimension reduction problems. A low-dimensional
projection represented as a scatterplot, binned plot, or contour plot,
will still suffer from data piling and reduced resolution near the
center of the projection. These approaches are still useful for large
sample sizes, and could be applied after the transformation.

The underlying idea of the (radial) transformation, to reverse the
curse, is to correct for the distortion in geometry that occurs when
projecting from a high-dimensional space onto low dimensions. This is
done by mapping equal volume in the full space onto equal area in a
two-dimensional projection via a geometrical transformation of the
projected data. When used with data that is uniformly distributed in the
high-dimensional space, the transformation will result in a uniform
distribution in the low-dimensional display. In general, for
\textit{any} underlying data distribution, we can instead think about
the effect in terms of resolution: adjusting for the geometric piling
effect means that the resolution depends on the projected volume. The
observed projected distribution is representative of the original
high-dimensional distribution in this sense.

@@

To understand how the \(p\)-dimensional volume is projected onto a
\(2\)-dimensional plane, we study what fraction of the total volume is
projected onto the area of a disk depending on its radius. This
dependence was described in \citet{Laa:2020wkm}. We start from a
\(p\)-dimensional hypersphere, with radius \(R\) and volume \(V(R, p)\),
and its projected volume onto a centered \(2\)-dimensional disk of
radius \(r\), \(V_{2D}(r; p, R)\), where \(r\) can be any radius within
\([0, R]\). The relative projected volume is then given as the ratio of
these two quantities, \begin{equation}
v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation}

@@

\textcolor{red}{We transform $r_y$ in two steps. First, we replace it with $v_{2} (r_y; p, R)$ to redistribute the volume uniformly by the radius. The second step is then to }
transform \(v_{2} (r_y; p, R)\) using the inverse of
\(v_{2} (r_y; 2, R)\), to go from a uniform distribution in radius to a
uniform distribution in area of the disk. This inverse is defined via
\(v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y\)
and thus \(v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}\).

@@

\caption{Concentric circles that are equidistant before the application of the radial transformation, for $p=2, 3, 10, 100$, illustrating the effect of the transformation with increasing number of dimensions. The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.}\label{fig:circles}
\end{figure}

\textcolor{red}{At this point a few characteristics of the transformation should be noted. First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph. In addition, since only the radial component is transformed, angles between the points will not be altered. (This can also be shown explicitly, ). Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$). For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space. Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.}

@@

Before applying the transformation, it can be important to preprocess
the data, as is generally required with methods for the visualisation of
high-dimensional data. In particular, when working with projections that
combine multiple variables it is important to ensure a common scale
across the dimensions. This is especially important with the sage
display, where the overall scale (set by the maximum radius \(R\))
determines the effect of the transformation. Scaling is commonly also
combined with centering as a preprocessing step. This is less important
in our implementation because we always center the projected data before
applying our transformation (i.e.~the projected mean defines the origin
of our polar coordinate system in the plane). Centering and scaling
together would be achieved by standardizing each variable in the data to
be centered at 0 and have a variance equal to 1.

We may also wish to apply non-linear transformations before visualising
the data or combining variables in a projection. One example would be to
use a log transformation in the case of a variable that has a skewed
distribution. These types of transformations will generally change the
distribution across the high-dimensional volume (and in particular also
the position of the mean), and can thus make the sage display more
useful for a wide range of observed data distributions.

The third common pre-processing would involve a dimension reduction
step. Even with a relatively low number of dimensions, e.g.~with
\(p=10\), we see noticeable piling. However, we are often confronted
with much higher dimensional datasets in which many of the variables are
primarily adding noise. In this case it is advisable to first reduce the
number of dimensions, for example using principal components analysis.
Other dimension reduction techniques could be used. The sage tour would
then be used to view the first few components, which would likely
provide more insight about the data structures than is possible from
only the first two.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tourr}\SpecialCharTok{::}\FunctionTok{animate}\NormalTok{(}
\NormalTok{  data,}
  \AttributeTok{tour\_path =}\NormalTok{ tourr}\SpecialCharTok{::}\FunctionTok{grand\_tour}\NormalTok{(),}
  \AttributeTok{display =} \FunctionTok{display\_sage}\NormalTok{(gam, R, half\_range)}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

\noindent and uses \texttt{gam} to set the \(\gamma\) parameter for
computing \(p_{\mathrm{eff}}\), and the overall range \texttt{R} used
for trimming. Both these parameters are described in Section
\ref{sec:params} above.

For consistency with other display functions in the \texttt{tourr}
package we also pass in the scale parameter \(s\) via the
\texttt{half\_range} argument. By default this is evaluated from the
range of the data, but kept as a free parameter to zoom in (without any
transformation) if necessary. This can be important, e.g.~when the range
is large because of an outlying point. Note here that the ratio \(R/s\)
sets the scale for fitting the displayed data on the plotting canvas, by
default \(s = R\). When adjusting \(R\) the user should take care to
adjust \(s\) accordingly.

@@

An alternative is the slice tour \citep{sliceTour} which allows
distributions of points around the center of the data to be explored
using sections instead of projections. Instead of adjusting for effects
from projecting a high-dimensional volume, the solution with the slice
tour is to only highlight projected points if they are within a small
volume around the current projection plane, thus showing a local view of
the distribution. The slice tour is useful when there are large numbers
of observations or if there is concave structure in the data. Note that
only few methods are currently available for looking at sections of the
data in a systematic fashion, and further work is needed to fully
explore the potential of this type of visualisation. In smaller datasets
we may also use interactive approaches such as linked brushing to
explore for local structures or conditional features.
